Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
M & M intro moved back to Intro

\
\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in terms of hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.

\
\   We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

Table \@ref(tab:hypo) here.

# Materials and Methods
\
\      We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We defined species ERD as the depth at which soil water potential best explained species-specific 5-yrly mean growth rates over 25 years. To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model. The ERD model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD. We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets. Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.

\
Fig. \@ref(fig:schematic) here.

In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Details of all datasets used are described in the Supplementary Information.

>>>>

\
\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in terms of hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

\
\      We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We defined species ERD as the depth at which soil water potential best explained species-specific 5-yrly mean growth rates over 25 years. To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model. The ERD model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD. We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets. Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.  Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

\
Fig. \@ref(fig:schematic) here.
Table \@ref(tab:hypo) here.

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Details of all datasets used are described in the Supplementary Information.
